Figure 3 - ChIP

### 2024-07-09 ms_figure_3_chip.Rmd
further updates to figure 3

---
title: "Figure 3 - ChIP"
author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    genomefasta: "reference/Schizosaccharomyces_pombe.ASM294v2.dna.toplevel.fa"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    chiptrackfile: "data/genome_browser_tracks_chipseq.csv.gz"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    dbdtxt: "data/TF_DBD_annotation.txt"
    motifrds: "data/motifs_annotated.rds"
    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
editor_options: 
    chunk_output_type: console
---

@@

@@

@@

@@

hm_prom_bound <- grid.grabExpr(
    {
        hm6 <- draw(hm6, merge_legend = TRUE,
                    heatmap_legend_side = "bottom",
                    annotation_legend_side = "bottom")
        decorate_heatmap_body("hm6", {
            grid.lines(c(mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)])), 0),
                       c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
        }, slice = 1)
        # decorate_heatmap_body("hm6", {
        #     grid.lines(c(1, mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)]))),
        #                c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
        # }, slice = 2)
    },
    width = w_peaksHm, height = w_peaksHm * mean(keep)
)
plot_grid(hm_prom_bound)
```

@@

@@

Assemble the panels into Figure 2:

```{r Fig2}
#| fig.width: 11.5
#| fig.height: 8.2
#| warning: false

cowplot::plot_grid(
    hm_prom,
    cowplot::plot_grid(gnmtracks1, scale = 0.9),
    hm_motifs_small,
    cowplot::plot_grid(
        motifs_aligned, upsets_aligned,
        nrow = 1, rel_widths = c(1.25, 1)),
    nrow = 2,
    labels = c("A", "B", "C", "D"),
    align = "vh",
    axis = "b",
    rel_widths = c(1, 1),
    rel_heights = c(w_peaksHm + 10, w_peaksHm + 20)
)
```

@@

Assemble the panels into Supplementary Figure 2:

```{r suppFig2}
#| fig.width: 11
#| fig.height: 9
#| warning: false

cowplot::plot_grid(
    cowplot::plot_grid(NULL, NULL, NULL, # space for labels
                       nrow = 1, labels = c("A", "B", "C"),
                       rel_widths = c(130, 100, 100)),
    cowplot::plot_grid(
        hm_ipms_hot,
        cowplot::plot_grid(gnmtracks2, NULL, ncol = 1, rel_heights = c(12, 1)),
        p_fracbound,
        nrow = 1, 
        labels = c("", "", ""),
        align = "v",
        axis = "b",
        rel_widths = c(130, 100, 100),
        rel_heights = c(180)
    ),
    nrow = 2,
    rel_widths = c(1),
    rel_heights = c(0.25, 9.75)
)
```

>>>>

---
title: "Figure 3 - ChIP"
author: "Michael Stadler, Charlotte Soneson, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    genomefasta: "reference/Schizosaccharomyces_pombe.ASM294v2.dna.toplevel.fa"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    chiptrackfile: "data/genome_browser_tracks_chipseq.csv.gz"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    dbdtxt: "data/TF_DBD_annotation.txt"
    motifrds: "data/motifs_annotated.rds"
    rdsrnaseq_del: "data/rnaseq_del.rds"
    txcsv: "data/promoters-1kb_tes-1kb_annotated.csv.gz"
editor_options: 
    chunk_output_type: console
---

@@

# make first letter of a string uppercase
.capitalize <- function(x) {
    paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}

@@

# `rnaseq_del`: a `SummarizedExperiment` with RNA-seq quantifications
rnaseq_del <- readRDS(params$rdsrnaseq_del)

@@

We create the same heatmap in long format, allowing to label all rows:

```{r}
#| fig.width: 6
#| fig.height: 8

hm4_long <- Heatmap(matrix = motifsim2, name = "Motif\nsimilarity",
               col = circlize::colorRamp2(breaks = seq(rng[1], rng[2], length.out = 64),
                                          colors = cols),
               cluster_rows = motifcl,
               show_row_dend = TRUE, row_dend_width = unit(10, "mm"),
               cluster_columns = motifcl, show_column_dend = FALSE,
               row_split = n_split, column_split = n_split,
               column_title = " ",
               row_gap = unit(0.2, "mm"), column_gap = unit(0.2, "mm"),
               show_row_names = TRUE, row_names_side = "right",
               column_title_gp = gpar(fontsize = fl),
               row_title = " ",
               row_names_gp = gpar(fontsize = fs),
               show_column_names = FALSE,
               heatmap_legend_param = list(
                   at = round(c(rng[1], mean(rng), rng[2]), 2),
                   labels_gp = gpar(fontsize = fl),
                   border = "gray10",
                   legend_direction = "horizontal",
                   legend_width = unit(18, "mm"),
                   title_position = "lefttop",
                   title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm * 0.5, "mm"), # heatmap body width
               height = unit(h_peaksHm * 1.15, "mm")) # heatmap body height

set.seed(42L)
hm_motifs_long <- grid.grabExpr(
    hm4_long <- draw(hm4_long, merge_legend = FALSE,
                     heatmap_legend_side = "bottom",
                     annotation_legend_side = "bottom"),
    width = w_peaksHm * 0.5, height = h_peaksHm * 1.15
)

plot_grid(hm_motifs_long)
```

@@

# manually create a legend
hm_legend <- Legend(labels = c("TFs binding to another TF promoter",
                               "TFs binding to their own promoter"),
                    title = " ", legend_gp = gpar(fill = c("#648FFF", "darkblue")))

hm_prom_bound <- grid.grabExpr({
        hm6 <- draw(hm6,
                    annotation_legend_side = "bottom",
                    annotation_legend_list = packLegend(hm_legend))
        decorate_heatmap_body("hm6", {
            grid.lines(c(mean(grepl("^Bound", promtype[match(tfnms, tfnms_all)])), 0),
                       c(0, 1), gp = gpar(lty = 2, lwd = 0.5))
        }, slice = 1)
    },
    width = w_peaksHm, height = w_peaksHm * mean(keep)
)
plot_grid(hm_prom_bound)
```

@@

### Network around `Atf1`

For clearer visualization, we create a sub-network including only the
connectsion to and from `Atf1`:

```{r make-network-atf1}
#| fig.width: 10
#| fig.height: 8

## Create graph
## ... identify TFs connected to atf1
to_keep <- rownames(adjm)[adjm["atf1", ] > 0 | adjm[, "atf1"] > 0]
length(to_keep)

## ... create adjacency matrix (row: from TF, column: to TF promoter)
adjm2 <- adjm[to_keep, to_keep]
gr2 <- graph_from_adjacency_matrix(adjm2, mode = "directed")

## Set attributes for vertices and edges
V(gr2)$vertexclass <- ifelse(adjm[cbind(names(V(gr2)), names(V(gr2)))] > 0,
                             "Auto-regulated TF",
                             ifelse(names(V(gr2)) %in% colnames(mprom),
                                    "ChIP'ed TF", "Other TF"))
V(gr2)$shape <- "circle"
eL <- as_edgelist(gr2)
recipr <- paste0(eL[,1], "-", eL[,2]) %in% paste0(eL[,2], "-", eL[,1]) & eL[,1] != eL[,2]
E(gr2)$weight <- ifelse(recipr, 2, 1)
E(gr2)$edgeclass <- ifelse(recipr, "Reciprocal", "Other")
E(gr2)$color <- ifelse(recipr, gplots::col2hex(main_colors[3]), gplots::col2hex(na_color))

## Visualize
node_padding <- margin(.2, .2, .2, .2, "mm")
ggr2 <- ggraph(as_tbl_graph(gr2), 
               layout = "igraph", algorithm = "fr") + 
    geom_edge_link(aes(start_cap = label_rect(node1.name, padding = node_padding),
                       end_cap = label_rect(node2.name, padding = node_padding),
                       color = edgeclass),
                   arrow = arrow(length = unit(2, 'mm'))) +
    geom_edge_loop(aes(start_cap = label_rect(node1.name, padding = node_padding),
                       end_cap = label_rect(node2.name, padding = node_padding)),
                   arrow = arrow(length = unit(2, 'mm'))) +  # Handle self-loops
    scale_edge_color_manual(values = c(Reciprocal = main_colors[3], 
                                       Other = gplots::col2hex(na_color)),
                            name = "") + 
    geom_node_point(size = 4, aes(fill = vertexclass, shape = vertexclass), 
                    color = "black", stroke = 0.1) + 
    scale_shape_manual(values = c("Other TF" = 21, "ChIP'ed TF" = 21, "Auto-regulated TF" = 22), name = "") +
    scale_fill_manual(values = c("Other TF" = gplots::col2hex(bg_color),
                                 "ChIP'ed TF" = gplots::col2hex(main_colors[5]),
                                 "Auto-regulated TF" = main_colors[5]),
                      name = "") + 
    geom_node_text(aes(label = name), size = 3.5, repel = TRUE, vjust = 1.8) + 
    theme_graph(base_family = "sans", plot_margin = margin(30, 10, 40, 10)) + 
    guides(fill = guide_legend(override.aes = list(size = 4, alpha = 1))) + 
    coord_cartesian(clip = "off") 
print(ggr2)
```

@@

Assemble the panels into Figure 3:

```{r Fig3}
#| fig.width: 11.5
#| fig.height: 8.2
#| warning: false

cowplot::plot_grid(
    hm_prom_bound,
    cowplot::plot_grid(gnmtracks1, scale = 0.9),
    hm_motifs_small,
    cowplot::plot_grid(
        motifs_aligned, upsets_aligned,
        nrow = 1, rel_widths = c(1.25, 1)),
    nrow = 2,
    labels = c("A", "B", "C", "D"),
    align = "vh",
    axis = "b",
    rel_widths = c(1, 1),
    rel_heights = c(w_peaksHm + 10, w_peaksHm + 20)
)
```

@@

Assemble the panels into Supplementary Figure 3:

```{r suppFig3}
#| fig.width: 11
#| fig.height: 9
#| warning: false

cowplot::plot_grid(
    cowplot::plot_grid(NULL, NULL, NULL, # space for labels
                       nrow = 1, labels = c("A", "B", "C"),
                       rel_widths = c(100, 100, 100)),
    cowplot::plot_grid(
        cowplot::plot_grid(gnmtracks2, NULL, ncol = 1, rel_heights = c(20, 1)),
        hm_motifs_long,
        p_fracbound,
        nrow = 1,
        labels = c("", "", "")
    ),
    nrow = 2,
    rel_widths = c(1),
    rel_heights = c(0.25, 9.75)
)
```
